\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-03-05 analysis/paper/paper.Rmd
In process of changing my results
loaded data, now ready to process

For the third case study, we will look at regularized estimation for the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. Before imputation, there were a total of 9105 individuals and 47 variables. Of those variables, a few were removed before imputation. We removed three response variables: hospital charges, patient ratio of costs to charges and patient micro-costs. We also removed functional disability and income as they are ordinal covariates, because ordinal variables are treated as factors in our current model. Finally, we removed 8 covariates relating to the results of previous findings: we removed SUPPORT day 3 physiology score (sps), 	APACHE III day 3 physiology score (aps),	SUPPORT model 2-month survival estimate, SUPPORT model 6-month survival estimate, Physician's 2-month survival estimate for pt., Physician's 6-month survival estimate for pt., Patient had Do Not Resuscitate (DNR) order and Day of DNR order (<0 if before study). Of these, sps and aps were added on after imputation, as they were missing only 1 observation and permitted us to compare findings between our model and their's.

@@

hospdead      & Death in Hospital                        &                 & double           & 0            \\

@@

charges       & Hospital Charges                         &                 & double           & 25           \\
totcst        & Total RCC cost                           &                 & double           & 105          \\
totmcst       & Total micro-cost                         &                 & double           & 372          \\

@@

adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
sfdm2         &                                          & 5               & integer          & 159          \\

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}.

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}
data(support)
support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

Before we proceed with the analysis, we first remove hospital death as a covariate, as this is directly informative of death. All covariates will be used to estimate the hazard function. The hazard function will be used to estimate the cumulative incidence of each individual in the study. Then, the average cumulative incidience curve for all observations will be used for our comparison.

```{r supportSetUp, echo = FALSE, eval = eval_cs3}
completeData=casebase::support
#remove hospital dead
completeData<-completeData[,-c(4)]
ratio=10  #universal ratio for eval_cs3
#keep one individual out, at random, for which we will develop an 
#absolute risk curve.
#sam=sample(1:nrow(completeData), 1)
#Set workingData to the remaining individuals in the data
#workingData=completeData[-c(sam),]
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])
```

@@

```{r cbHazardInteractions}, echo = TRUE, eval = eval_cs3}
xInteractions=x*y[,2]
colnames(xInteractions) <- paste("d.time*", colnames(xInteractions), sep = "")
xFull=cbind(x,xInteractions)
cb.ModelInter=casebase::fitSmoothHazard.fit(xFull,y,family="glmnet",time="d.time",event="death",
                                       formula_time = ~d.time,alpha=1,ratio=rat)
```
Then, we take our model and use the absoluteRisk function to integrate and retrieve our absolute risk function.

```{r CBabsoluteInteracts, echo = TRUE, eval = eval_cs3}
#Estimating the absolute risk curve using the newData parameter.
casebaseAbsoluteInter=casebase::absoluteRisk(cb.ModelInter,time = seq(1,max(completeData$d.time), 400),newdata = xFull , s="lambda.1se",method=c("numerical"))

cabI=data.frame(casebaseAbsoluteInter[,1],rowMeans(casebaseAbsoluteInter[,-c(1)]))
```
We will compare this curve to a regularized version of cox regression, and a kaplan-meier survival curve. The cox regression absolute risk curve with lasso penalization required two extra steps to retrieve. Coxnet from the glmnet package has tools to fit a regularized cox model, but requires a survival object skeleton of the selected variables, so that the survival package tools can be used to retrive the absolute risk. We handle this by fitting a regularized cox model with the glmnet package, and a cox model from the survival package with the selected variables from the regularized fit. This second model serves as a skeleton that permits the use of survival package functions. The coefficients from the regularized model will replace the coefficients in our skeleton. the resulting model will have the correct coefficients when integrating, but will have an incorrect standard error. For the purposes of this case study, we are only interested in the absolute risk curve itself and not the standard error.

```{r coxHazAbsoluteInteracts, echo = TRUE, eval = eval_cs3}
#Create u and xCox, which will be used when fitting Cox with glmnet.
u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))

xCoxI=as.matrix(cbind(data.frame("(Intercept)"=rep(1,each=690)),xFull))
colnames(xCoxI)[1]="(Intercept)"

#hazard for cox using glmnet
coxNetI=glmnet::cv.glmnet(x=xCoxI,y=u, family="cox",alpha=1)
#convergence demonstrated in plot
#plot(coxNet)
#taking the coefficient estimates for later use
nonzero_covariate_cox <- predict(coxNetI, type = "nonzero", s = "lambda.1se")
nonzero_coef_cox <- coef(coxNetI, s = "lambda.1se")
#creating a new dataset that only contains the covariates chosen through glmnet.
#cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
cleanCoxDataI<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,nonzero_covariate_cox$X1]))
#newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
#fitting a cox model using regular estimation, however we will not keep it.
#this is used more as an object place holder.
coxNetFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxDataI)
#The coefficients of this object will be replaced with the estimates from coxNet.
#Doing so makes it so that everything is invalid aside from the coefficients.
#In this case, all we need to estimate the absolute risk is the coefficients.
#Std. error would be incorrect here, if we were to draw error bars.
coxNetFitI$coefficients<-nonzero_coef_cox@x
#Fitting absolute risk curve for cox+glmnet. -1 to remove intercept
#posCox=nonzero_covariate_cox$X1
#newDataCoxI=cleanCoxDataI[,posCox]

abcoxNetI<-survival::survfit(coxNetFitI,type="breslow",newdata=cleanCoxDataI)

abCNI<-abcoxNetI
abCNI$surv=rowMeans(abCNI$surv)
```

```{r unpenalizedCox, echo = TRUE, eval = eval_cs3}
fullDataCox<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCoxI[,-c(1)]))
coxFitI <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = fullDataCox)
abcoxFitI<-survival::survfit(coxFitI,newdata=fullDataCox)
abCFI<-abcoxFitI
abCFI$surv<-rowMeans(abCFI$surv)
#cb.ModelNorm=casebase::fitSmoothHazard(V2~.,time="V1",event="V2",
       #                                ratio=rat,data=fullDataCox)
#summary(cb.ModelNorm)#need to figure out how to deal with NAs
#casebaseAbsoluteNorm=casebase::absoluteRisk(cb.ModelNorm,time = seq(1,max(completeData$d.time), 200),newdata = newData , s="lambda.1se",method=c("numerical"))
#cabn=data.frame(casebaseAbsoluteNorm[,1],rowMeans(casebaseAbsoluteNorm[,-c(1)]))

```
We used the survival package to calculate the kaplan-meier absolute risk function.

```{r fitKM+abRisk, echo = TRUE, eval = eval_cs3}

coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)

#creating a surv object to be used to fit an unadjusted absolute risk curve.
# (Kaplan-Meier risk curve)
km <- survival::Surv(time = completeData$d.time, event = completeData$death)
abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
```

Now that our four cumulative incidence estimations have been calculated, we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase decrease the absolute risk by the end of the study, in comparison to kaplan-meier.
```{r abSupportComparison, echo = FALSE, eval = eval_cs3, fig.cap="\\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue."}
abCFup<-abcoxFitup
abCFup$surv<-rowMeans(abCFup$surv)
# A plot to compare all three Absolute risk (Commulative Incidence)
plot(cabI[,1],cabI[,2],type='l',col="black",lwd=3,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
lines(abCFI,col="red",fun="event",lwd=3,conf.int = FALSE)
lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
lines(abCNI,col="purple",fun="event",lwd=3,conf.int = FALSE)
legend("bottomright", 
       legend = c( "Casebase (Lasso+linear)","semi-parametric Coxnet","KM curve","cox"), 
       col = c("black","red","Blue","purple"),
       lty = c(1, 1, 1), 
       bg = "gray90")
```

```{r covariatesRegularizedLollipopPlot,eval=FALSE}

#get covariates excluding d.time
estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxNet)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimates=rbind(estimatescb, estimatescox)
estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
        geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
        geom_point(position=ggstance::position_dodgev(height=0.2))

```

```{r covariatesLolliInteractions,eval=eval_cs3}
#get covariates excluding d.time
library(data.table)
estimatescb=setDT(as.data.frame(coef(cb.ModelInter)[-c(1,2),1]), keep.rownames = TRUE)[]
estimatescb$Model="casebase"
colnames(estimatescb)<-c("Coefs","Estimate","Model")
estimatescox=setDT(as.data.frame(coef(coxNetI)[-c(1),1]), keep.rownames = TRUE)[]
estimatescox$Model="Cox"
colnames(estimatescox)<-c("Coefs","Estimate","Model")
estimatesI=rbind(estimatescb, estimatescox)
estimatesI$Estimate[estimatesI$Estimate==0]=NA# make it so unchosen covariates appear empty
library(ggstance)# for vertical dodge
  ggplot(estimatesI, aes(x = Coefs, y = Estimate , color = Model)) +
        geom_segment(aes(y = 0, x = Coefs, yend = Estimate, xend = Coefs),size=0.5, position = position_dodge(width = 0.5)) +
        geom_point(position = position_dodge(width = 0.5))

```

>>>>

For the third case study, we will look at regularized estimation for the hazard function. This can be performed using a combination of case-base sampling and regularized logistic regression. To demonstrate this capability, we examine the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset [@knaus1995support]. The SUPPORT dataset tracks death for individuals who are considered seriously ill within 5 hospitals in the United States of America. Imputation was conducted using the \pkg{mice} package in \proglang{R} with default settings [@mice]. Before imputation, there were a total of 9105 individuals and 47 variables. Of those variables, a few were removed before imputation. We removed three response variables: hospital charges, patient ratio of costs to charges and patient micro-costs. Next, we removed hospital death as it was directly informative of our event of interest, death. We also removed functional disability and income as they are ordinal covariates, because ordinal variables are treated as factors in our current model. Finally, we removed 8 covariates relating to the results of previous findings: we removed SUPPORT day 3 physiology score (sps), 	APACHE III day 3 physiology score (aps),	SUPPORT model 2-month survival estimate, SUPPORT model 6-month survival estimate, Physician's 2-month survival estimate for pt., Physician's 6-month survival estimate for pt., Patient had Do Not Resuscitate (DNR) order and Day of DNR order (<0 if before study). Of these, sps and aps were added on after imputation, as they were missing only 1 observation and permitted us to compare findings between our model and their's.

@@

@@

@@

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
ca       & Yes                                           \\
         & No                                            \\
         & Metastatic                                    \\
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

<!--Hospital death as a covariate was removed, as this is directly informative of death. The glmnet package on CRAN was used to introduce elastic net, lasso and ridge regressions to casebase.--> As before, the first step is to visualize the incidence density. The population-time plot for the SUPPORT data appears in Figure \ref{fig:support-poptime}. As we can see, the incidence density is not constant.

```{r support-poptime, echo = TRUE, eval = eval_cs3, fig.cap="\\label{fig:support-poptime} Population-time plot for the SUPPORT data."}

completeData<-casebase::support
workingCompleteData=model.matrix(death~ .-d.time,data=completeData)[,-c(1)]#remove intercept
#Create u and xCox, which will be used when fitting Cox with glmnet.
newData=workingCompleteData
x=workingCompleteData
#x and y will be used to fit the casebase model under glmnet.
#x=as.matrix(sparse.model.matrix(death~ .-d.time+0,data=workingData))
y=data.matrix(completeData[,c(2,5)])

support_popTime <- popTime(support,
                           time = "d.time", event = "death",percentile_number = 0.31)
plot(support_popTime)
```

Our next step is to get the hazard function. All covariates will be used to estimate the hazard function. The hazard function will be used to estimate the cumulative incidence of each individual in the study. Then, the average cumulative incidience curve for all observations will be used for our comparison.

@@
